# Appraising the causal relevance of DNA methylation for risk of lung cancer {#dnam-lung-cancer-mr}

```{r call-source-07}
read_chunk("chunks/07-dnam_lungcancer_mr.R")
```

``` {r load-data-07, results="hide", message=FALSE, warning=FALSE}
```

``` {r tables-setup-07, results="hide", message=FALSE, warning=FALSE}
```

``` {r figures-setup-07, results="hide", message=FALSE, warning=FALSE}
```


## Abstract {#abstract-07}

### Background {#abstract-background-07}
DNA methylation changes in peripheral blood have recently been identified in relation to lung cancer risk. Some of these changes have been suggested to mediate part of the effect of smoking on lung cancer. However, limitations with conventional mediation analyses mean that the causal nature of these methylation changes has yet to be fully elucidated.

### Methods {#abstract-methods-07}
I first performed a meta-analysis of four epigenome-wide association studies (EWAS) of lung cancer (918 cases, 918 controls). Next, I conducted a two-sample Mendelian randomization (MR) analysis, using genetic instruments for methylation at CpG sites identified in the EWAS meta-analysis, and 29,863 cases and 55,586 controls from the TRICL-ILCCO lung cancer consortium, to appraise the possible causal role of methylation at these sites on lung cancer.

### Results {#abstract-results-07}
Sixteen CpG sites were identified from the EWAS meta-analysis (FDR < 0.05), 14 of which had suitable genetic instruments. MR provided little evidence that DNA methylation in peripheral blood at the 14 CpG sites play a causal role in lung cancer development (FDR > 0.05), including for cg05575921-_AHRR_ where methylation is strongly associated with both smoke exposure and lung cancer risk.

### Conclusions {#abstract-conclusions-07}
The results contrast with previous observational and mediation analysis, which have made strong claims regarding the causal role of DNA methylation. Thus, previous suggestions of a mediating role of methylation at sites identified in peripheral blood, such as cg05575921-_AHRR_, could be unfounded. However, this study does not preclude the possibility that differential DNA methylation at other sites is causally involved in lung cancer development, especially within lung tissue.

## Introduction {#introduction-07}
Lung cancer is the most common cause of cancer-related death worldwide [@Ferlay2013]. Several DNA methylation changes have been recently identified in relation to lung cancer risk [@Fasanelli2015; @Baglietto2017; @McCarthy2016]. However, these epigenetic marks are sensitive to reverse causation, being affected by cancer processes [@Jones2002], and are also prone to confounding, for example by socio-economic and lifestyle factors [@Borghol2012; @Elliott2014].

One CpG site, cg05575921 within the aryl hydrocarbon receptor repressor (_AHRR_) gene, has been consistently replicated in relation to both smoking [@Joehanes2016] and lung cancer [@Fasanelli2015; @Baglietto2017; @Bojesen2017] and functional evidence suggests that this region could be causally involved in lung cancer [@Zudaire2008]. However, the observed association between methylation and lung cancer might simply reflect separate effects of smoking on lung cancer and DNA methylation, i.e. the association may be a result of confounding [@Richmond2016], including residual confounding after adjustment for self-reported smoking behaviour [@Fewell2007; @Munafo2012].  Furthermore, recent EWAS of lung cancer have revealed additional CpG sites which may be causally implicated in development of the disease [@Fasanelli2015; @Baglietto2017]. 

As discussed in __Section \@ref(establishing-causality)__, MR can be used to help infer causality in associations between DNA methylation and complex traits [@Relton2012; @Relton2015; @Richardson2017].

Mendelian randomization (MR) uses genetic variants associated with modifiable factors as instruments to infer causality between the modifiable factor and outcome, overcoming most unmeasured or residual confounding and reverse causation [@DaveySmith2014; @DaveySmith2003]. In order to infer causality, three core assumptions of MR should be met: 1) The instrument is associated with the exposure, 2) The instrument is not associated with any confounders, 3) The instrument is associated with the outcome only through the exposure. MR may be adapted to the setting of DNA methylation [@Relton2012; @Relton2015; @Richardson2017] with the use of single nucleotide polymorphisms (SNPs) that correlate with methylation of CpG sites, known as methylation quantitative trait loci (mQTLs) [@Gaunt2016].

In this study, I performed a meta-analysis of four lung cancer EWAS (918 case-control pairs) from prospective cohort studies to identify CpG sites associated with lung cancer risk and applied MR to investigate whether the observed DNA methylation changes at these sites are causally linked to lung cancer.

In this chapter, Dr Rebecca Richmond performed analysis in the CCHS cohort (see __Sections \@ref(ahrr-one-sample-mr-methods) and \@ref(ahrr-one-sample-mr)__) and contributed to writing the introduction and discussion. Dr Andrew E. Teschendorff analysed differences in lung tumour and adjacent healthy tissue (see __Sections \@ref(lc-heathly-v-normal-methods) and \@ref(lc-heathly-v-normal)__). I completed the rest of the analyses and wrote each section, except those for which Dr Rebecca Richmond completed the analyses.

## Methods {#methods-07}

### EWAS study details
A meta-analysis of four lung cancer case-control EWAS was conducted to identify DNA methlyation sites associated with lung cancer. DNA methylation in each EWAS was assessed using the Illumina Infinium® HumanMethylation450 (HM450) BeadChip. All EWAS are nested within prospective cohorts that measured DNA methylation in peripheral blood samples before diagnosis: EPIC-Italy (185 case-control pairs), Melbourne Collaborative Cohort Study (MCCS) (367 case-control pairs), Norwegian Women and Cancer (NOWAC) (132 case-control pairs) and the Northern Sweden Health and Disease Study (NSHDS) (234 case-control pairs). Study populations, laboratory methods, data pre-processing and quality control methods have been described in detail elsewhere [@Baglietto2017] and are outlined below.

At the various laboratory sites, samples were distributed into 96-well plates and processed in chips of 12 arrays (8 chips per plate) with case-control pairs arranged randomly on the same chip. Methylation data were pre-processed and normalized in each study, and probe filtering was performed as previously described [@Baglietto2017], leaving 465,886 CpGs suitable for the analysis in EPIC-Italy, 485,330 CpGs in MCCS, 450,890 CpGs in NOWAC and 482,867 CpGs in NSHDS.

#### European Prospective Investigation into Cancer and Nutrition-Italy (EPIC-Italy)
EPIC-Italy includes 47,749 volunteers (32,579 women) aged 35–70 years at the time of recruitment (1992–1998). Anthropometric measurements and lifestyle variables including detailed information on smoking history were collected at recruitment through standardized questionnaires, together with a blood sample. Within EPIC-Italy a nested case-control study was conducted utilizing incident cases diagnosed within follow-up and healthy controls individually matched to cases by gender, date of birth (±5 years), date of inclusion in the study and study centre. Analysis was performed for 185 incident cases diagnosed within follow-up and matched controls. Laboratory procedures were carried out at the Human Genetics Foundation (Turin, Italy) and DNA extracted from buffy coats as previously described [@Baglietto2017]. All participants signed an informed consent form, and the ethical review boards of the International Agency for Research on Cancer and of each local participating centre approved the study protocol.

#### Melbourne Collaborative Cohort Study (MCCS)
The MCCS is a prospective cohort study of 41,514 volunteers (24,469 women) aged between 27 and 76 years at baseline (1990-1994). At baseline attendance, participants completed questionnaires that measured demographic characteristics and lifestyle factors. Height and weight were directly measured, and a blood sample was collected and stored. Incident cases of lung cancer were identified through linkage with the State and National Cancer Registries during follow-up up to the end of 2011. The MCCS sample included 367 cases and 367 matched controls selected from MCCS participants who were lung cancer free at the age of diagnosis of the matching case (density sampling). Matching variables included gender, date of blood collection (within 6 months), date of birth (within 1 year), country of birth (Australia and UK versus Southern Europe), type of biospecimen (lymphocyte, buffy coat and dried blood spot) and smoking status (never smokers; short-term former smokers: quitting smoking less than 10 years before blood draw; long-term former smokers: quitting smoking 10 years or more before blood draw; current light smokers: less than 15 cigarettes per day at blood draw; and current heavy smokers: 15 cigarettes or more at blood draw). For the MCCS, laboratory procedures were carried out at the Genetic Epidemiology Laboratory, the University of Melbourne according to manufacturers' protocols. DNA extraction from lymphocytes and buffy coats was performed as previously described [@Baglietto2017]. The Cancer Council Victoria's Human Research Ethics Committee approved the study protocol. Subjects gave written consent to participate and for the investigators to obtain access to their medical records.

#### Norwegian Women and Cancer (NOWAC)
The biobank of the NOWAC cohort was established in the years 2003-2006. Those who filled in an eight-page questionnaire and accepted the invitation to donate blood were sent blood drawing equipment together with a two-page epidemiological questionnaire. Around 50 000 women returned two tubes of blood to the Institute of Community Medicine at UiT The Arctic University of Norway and data linkage to the National Cancer Registry of Norway was performed. During follow-up to the end of 2011, 132 eligible cases of lung cancer were identified and were used for the EWAS. For each case, one control with an available blood sample was selected and matched on time since blood sampling and year of birth in order to control for effects of storage time and ageing. The cases and the controls were processed together for all laboratory procedures in order to reduce any batch effect. Laboratory procedures were carried out at the Human Genetics Foundation (Turin, Italy). DNA extraction from buffy coats was performed as previously described [@Baglietto2017]. All participants gave informed consent. The study was approved by the Regional Committee for Medical and Health Research Ethics in North Norway. Data storage and linkage was approved by the Norwegian Data Inspectorate.

#### Northern Sweden Health and Disease Study (NSHDS)
NSHDS is an ongoing prospective cohort and intervention study intended for health promotion of the population of Västerbotten County in northern Sweden. All residents were invited to participate by attending a health check-up at their local health care centre at 40, 50 and 60 years of age. At the health check-up, participants were asked to complete a self-administered questionnaire covering various factors such as education, smoking habits, physical activity and diet. In addition, height and weight were measured and participants were asked to donate a blood sample. Incident lung cancer cases were identified through linkage to the regional cancer registry. One control was chosen at random for each lung cancer case from appropriate risk sets consisting of all cohort members alive and free of cancer (except non-melanoma skin cancer) at the time of diagnosis of the index case. Matching criteria were the same as for the MCCS except there was no matching for type of biospecimens as DNA was extracted from whole blood for all samples. After quality control, a total of 234 incident lung cancer cases and 234 individually matched controls were available for this analysis. Laboratory procedures for NSHDS were carried out at two sites. DNA extraction from the buffy coat was conducted at Umeå University, Sweden, as previously described. Illumina Infinium HumanMethylation450 BeadChip analysis was conducted at the ALSPAC/IEU Laboratory at the University of Bristol. All study subjects provided written informed consent at time of the recruitment into the NSHDS.

### EWAS Meta-analysis {#methods-ewas-meta-analysis}
To quantify the association between the methylation level at each CpG and the risk of lung cancer conditional logistic regression models were fitted for beta values of methylation (which ranges from 0 (no cytosines methylated) to 1 (all cytosines methylated)) on lung cancer status for the four studies. The cases and controls in each study were matched, details can be found above. Surrogate variables were computed in the four studies using the SVA R package [@Leek2016] and the proportion of CD8+ and CD4+ T cells, B cells, monocytes, natural killer cells and granulocytes within whole blood were derived from DNA methylation [@Houseman2012]. The following EWAS models were included in the meta-analysis: Model 1 – unadjusted; Model 2 – adjusted for 10 surrogate variables (SVs); Model 3 – adjusted for 10 SVs and derived cell proportions. EWAS stratified by smoking status was also conducted (never (N=304), former (N=648) and current smoking (N=857)). For Model 1 and Model 2, the case-control studies not matched on smoking status (EPIC-Italy and NOWAC) were adjusted for smoking.

An inverse-variance weighted fixed effects meta-analysis was performed of the EWAS (918 case-control pairs) using the [METAL software](http://csg.sph.umich.edu/abecasis/metal/). Direction of effect, effect estimates and the I^2^ statistic were used to assess heterogeneity across the studies in addition to effect estimates across smoking strata (never, former and current). All sites identified at a false discovery rate (FDR) < 0.05 in Model 2 and 3 were also present in the sites identified in Model 1. The effect size differences between models for all sites identified in Model 1 were assessed by a Kruskal-Wallis test and a post-hoc Dunn’s test. There was little evidence for a difference (P > 0.1), so to maximize inclusion into the MR analyses sites identified in the unadjusted model (Model 1) were taken forward.

### Mendelian randomization {#methods-mendelian-randomization-07}
Two-sample MR was used to establish potential causal effect of differential methylation on lung cancer risk [@Inoue2010; @Pierce2013].

#### Sample 1: Accessible Resource for Integrated Epigenomic Studies (ARIES)
In the first sample, mQTL-methylation effect estimates ($\beta_{GP}$) for each CpG site of interest were identified in an mQTL database from the Accessible Resource for Integrated Epigenomic Studies (ARIES) (http://www.mqtldb.org). Details on the methylation pre-processing, genotyping and quality control (QC) pipelines are described in __Section \@ref(alspac-03)__. 

#### Sample 2: Transdisciplinary Research in Cancer of the Lung and The International Lung Cancer Consortium
In the second sample, summary data was extracted from a GWAS meta-analysis of lung cancer risk conducted by the Transdisciplinary Research in Cancer of the Lung and The International Lung Cancer Consortium (TRICL-ILCCO) (29,863 cases, 55,586 controls) for individuals genotyped using the Illumina Infinium OncoArray-500K BeadChip (Illumina Inc. San Diego, CA) and independent samples for which prior genotyping was performed [@McKay2017]. They were used to obtain mQTL-lung cancer estimates ($\beta_{GD}$).

For each independent mQTL (r^2^ < 0.01), I calculated the log odds ratio (OR) per SD unit increase in methylation by the formula $\frac{\beta_{GD}} {\beta_{GP}}$ (Wald ratio). Standard errors were approximated by the delta method [@Thomas2007]. Where multiple independent mQTLs were available for one CpG site, these were combined in a fixed effects meta-analysis after weighting each ratio estimate by the inverse variance of their associations with the outcome. Heterogeneity in Wald ratios across mQTLs was estimated using Cochran’s Q test, which can be used to indicate horizontal pleiotropy [@Bowden2015]. Differences between the observational and MR estimates were assessed using a Z-test for difference.

If there was evidence for an mQTL-CpG site association in ARIES in at least one time-point, it was assessed whether the mQTL replicated across time points in ARIES (FDR < 0.05, same direction of effect). Further, this association was re-analysed using linear regression of methylation on each genotyped SNP available in an independent cohort (NSHDS), using rvtests [@Zhan2016]. The same NSHDS samples on which DNA methylation was measured were genotyped using the Illumina Infinium OncoArray-500k BeadChip (Illumina Inc. San Diego, CA) and quality control parameters were applied under the recently published TRICL-ILCCO GWAS study on lung cancer [@McKay2017]. Genetic imputation was performed on these samples using the Haplotype Reference Consortium (HRC) Panel (release 1) [@McCarthy2016] through the Michigan Imputation Server [@Das2016]. Replicated mQTLs were included where possible to reduce the effect of winner’s curse using effect estimates from ARIES. The instrument strength of the mQTLs were assessed by the variance explained in methylation by each mQTL (r^2^) as well as the F-statistic in ARIES __Table \@ref(tab:sup-tab1-07)__. The power to detect the observational effect estimates in the two-sample MR analysis was assessed _a priori_, based on an alpha of 0.05, sample size of 29,863 cases and 55,586 controls (from TRICL-ILCCO) and calculated variance explained (r^2^). \linebreak

```{r, sup-tab1-07, message=FALSE, results="asis"}
```

\linebreak

MR analyses were also performed to investigate the impact of methylation on lung cancer subtypes in TRICL-ILCCO: adenocarcinoma (11,245 cases, 54,619 controls), small cell carcinoma (2791 cases, 20,580 controls), and squamous cell carcinoma (7704 cases, 54,763 controls). The association in never smokers (2303 cases, 6995 controls) and ever smokers (23,848 cases, 16,605 controls) [@McKay2017] was also assessed. Differences between the smoking subgroups were assessed using a Z-test for difference.

Next the association was tested between the identified mQTLs and four smoking behaviour traits which could confound the methylation-lung cancer association: number of cigarettes per day, smoking cessation rate, smoking initiation and age of smoking initiation using GWAS data from the Tobacco and Genetics (TAG) consortium (N=74,053) [@Furberg2010].

### Supplementary analyses {#methods-supplementary-analyses-07}

#### Assessing the potential causal effect of _AHRR_ methylation: one sample MR {#ahrr-one-sample-mr-methods}
This section (until the heading "Tumour and adjacent normal methylation patterns") was written by Dr Rebecca Richmond. 

Given previous findings implicating methylation at _AHRR_ in relation to lung cancer [@Fasanelli2015; @Baglietto2017], a one-sample MR analysis [@Haycock2016] was performed of _AHRR_ methylation on lung cancer incidence using individual-level data from the Copenhagen City Heart Study (CCHS) (357 incident cases, 8401 remaining free of lung cancer). Copenhagen City Heart Study is a prospective study of the general population [@Kaur-Knudsen2011]. Copenhagen residents were invited to complete a questionnaire and undergo a physical examination and are followed through a unique person identifier in the Danish health registries. All participants gave written informed consent, and a Danish ethics committee approved the study (KF100.2039/91).

#### Phenotypic data 
Participants were asked whether they smoked at the day of attendance or previously. If they answered affirmative to either of these questions, they were asked about their current and former smoking behaviour, including age of smoking initiation, age of smoking cessation, and number of daily consumed cigarettes, cheroots, cigars, and weekly grams of pipe tobacco. Based on these answers, participants were categorized as never, former, and current smokers. In addition, participants reported on alcohol consumption, occupational exposure to dust and/or welding fumes, exposure to passive smoking, education, and familial cases of lung cancer. The answers were reviewed together with an examiner at the day of attendance. Body mass index was calculated as measured weight in kilograms divided by measured height (in meters) squared.

#### Methylation data
At the physical examination, blood samples were drawn for DNA from which _AHRR_ methylation extent was measured [@Bojesen2017]. The _AHRR_ cg05575921 methylation extent was measured in duplicate samples of bisulphite treated DNA from peripheral blood from 9,234 individuals. A Taqman assay was used that was developed in the CCHS laboratory, and included standard curves, as well as internal controls in each 384-well plate. Coefficients of variation at the methylation level of 71% varied from 5.0 to 6.7%. Laboratory technicians were blinded to smoking and disease status of the individuals. Results were validated with pyrosequencing on a subset of samples.

#### Genetic data
Genotypes from the iCOGs array [@Bojesen2013] and prospective data on lung cancer incidence were also available for these participants. Of the 9234 individuals, genotype data from iCOGS on 8778 were available. In short, DNA isolated from leukocytes was genotyped with a custom Illumina iSelect genotyping array, designed to test genetic variants related to breast, ovary and prostate cancer, comprising roughly 211,000 SNPs after rigorous quality control [@Bojesen2013].

#### Lung cancer data
For lung cancer (ICD7, codes 1624 or 4624 until 1977, and ICD10, code C34 from 1978 and onwards), the date of first diagnosis was taken from the national Danish Cancer Registry from 1943 to December 2012.

#### Identification of mQTLs for CCHS one-sample MR
mQTLs located within 1Mb of cg05575921 _AHRR_ were identified in ARIES (FDR<0.05). Of those mQTLs which replicated within the CCHS, an LD pruning step was performed using a less stringent r^2^ threshold of 0.2 and generated an unweighted allele score, calculated by coding and then summing the alleles to reflect the average number of methylation-increasing alleles carried by an individual. Associations between the allele score and several potential confounding factors (sex, alcohol consumption, smoking status, occupational exposure to dust and/or welding fumes, passive smoking) were investigated. Then MR analyses were performed using two-stage Cox regression, with adjustment for age and sex, and further stratified by smoking status.

#### Tumour and adjacent normal methylation patterns {#lc-heathly-v-normal-methods}
This analysis was performed by Dr Andrew E. Teschendorff.

DNA methylation data from lung cancer tissue and matched normal adjacent tissue (N=40 squamous cell carcinoma and N=29 adenocarcinoma), profiled as part of The Cancer Genome Atlas (TCGA), were used to assess tissue-specific DNA methylation changes across sites identified in the meta-analysis of EWAS, as outlined previously [@Teschendorff2015].

#### mQTL association with gene expression
Gene expression at genes annotated to CpG sites identified in the lung cancer EWAS was examined in whole blood and lung tissue using data from the Gene-Tissue Expression (GTEx) consortium [@GTExConsortium2013].

Analyses were conducted in Stata (version 14) and R (version 3.2.2). For the two-sample MR analysis the MR-Base R package TwoSampleMR [@Hemani2016] was used. An adjusted P value that limited the FDR was calculated using the Benjamini-Hochberg method [@Benjamini1995]. All statistical tests were two-sided.

## Results {#results-07}
A flowchart representing the study design along with a summary of the results at each step is displayed in __Figure \@ref(fig:fig1-07)__.

(ref:fig1-07-cap) __Study design with results summary__. ARIES = Accessible Resource for Integrated Epigenomic Studies, TRICL-ILLCO = Transdisciplinary Research in Cancer of the Lung and The International Lung Cancer Consortium, MR = Mendelian randomization, CCHS = Copenhagen City Heart Study, TCGA = The Cancer Genome Atlas. * = 2000 individuals with samples at multiple timepoints.

(ref:fig1-07-scap) Study design with results summary

\blandscape

```{r fig1-07, message=FALSE, results="asis",fig.cap='(ref:fig1-07-cap)', fig.scap='(ref:fig1-07-scap)', fig.pos = "htbp"}
```

\elandscape

### EWAS meta-analysis {#results-ewas-meta-analysis}
The basic meta-analysis adjusted for study-specific covariates identified 16 CpG sites which were hypomethylated in relation to lung cancer (FDR<0.05, Model 1, __Figure \@ref(fig:fig2-07)__). Adjusting for 10 surrogate variables (Model 2) and derived cell counts (Model 3) gave similar results (__Table \@ref(tab:tab1-07)__). The direction of effect at the 16 sites did not vary between studies (median I^2^=38.6) (__Table \@ref(tab:sup-tab2-07)__), but there was evidence for heterogeneity of effect estimates at some sites when stratifying individuals by smoking status (__Table \@ref(tab:tab1-07)__).

(ref:fig2-07-cap) __Observational associations of DNA methylation and lung cancer: A fixed effects meta-analysis of lung cancer EWAS weighted on the inverse variance was performed to establish the observational association between differential DNA methylation and lung cancer__. Left-hand side: Manhattan plot, all points above the solid line are at P < 1x10^-7^ and all points above the dashed line (and triangular points) are at FDR < 0.05. In total 16 CpG sites are associated with lung cancer (FDR < 0.05). Right-hand side: Quantile-quantile plot of the EWAS results (same data as the Manhattan plot). 

(ref:fig2-07-scap) Observational associations of DNA methylation and lung cancer: A fixed effects meta-analysis of lung cancer EWAS weighted on the inverse variance was performed to establish the observational association between differential DNA methylation and lung cancer

\blandscape

```{r fig2-07, message=FALSE, results="asis", fig.show = "hold", out.width = "50%", fig.align = "default", fig.cap='(ref:fig2-07-cap)', fig.scap='(ref:fig2-07-scap)', fig.pos = "htbp"}
```

\elandscape

```{r tab1-07, message=FALSE, results="asis"}
```

```{r sup-tab2-07, message=FALSE, results="asis"}
```

### Mendelian randomization {#results-mendelian-randomization-07}
I identified 15 independent mQTLs (r^2^<0.01) associated with methylation at 14 of 16 CpGs. Ten mQTLs replicated at FDR<0.05 in NSHDS (__Table \@ref(tab:sup-tab3-07)__).  MR power analyses indicated >99% power to detect ORs for lung cancer of the same magnitude as those in the meta-analysis of EWAS.

```{r sup-tab3-07, message=FALSE, results="asis"}
```

There was little evidence for an effect of methylation at these 14 sites on lung cancer (FDR>0.05, __Table \@ref(tab:sup-tab4-07)__). For nine of 14 CpG sites the point estimates from the MR analysis were in the same direction as in the EWAS, but of a much smaller magnitude (Z-test for difference, P<0.001) (__Figure \@ref(fig:fig3-07)__).

```{r sup-tab4-07, message=FALSE, results="asis"}
```

\pagebreak

(ref:fig3-07-cap) __Mendelian randomization (MR) vs. observational analysis__. Two-sample MR was carried out with methylation at 14/16 CpG sites identified in the EWAS meta-analysis as the exposure and lung cancer as the outcome. cg01901332 and cg05575921 had 2 instruments so the estimate was calculated using the inverse variance weighted method, for the rest the MR estimate was calculated using a Wald ratio. Only 14 of 16 sites could be instrumented using mQTLs from mqtldb.org. * = instrumental variable not replicated in independent dataset (NSHDS). The sites for which instrumental variables have not been replicated are cg01901332, cg21566642, cg05575921 and cg08709672. OR = odds ratio per SD increase in DNA methylation.

(ref:fig3-07-scap) Mendelian randomization (MR) vs. observational analysis

```{r fig3-07, message=FALSE, results="asis", fig.cap='(ref:fig3-07-cap)', fig.scap='(ref:fig3-07-scap)'}
```

For 9 of out the 16 mQTL-CpG associations, there was strong replication across time points (__Table \@ref(tab:sup-tab5-07)__) and 10 out of 16 mQTL-CpG associations replicated at FDR<0.05 in an independent adult cohort (NSHDS).  Using mQTL effect estimates from NSHDS for the 10 CpG sites that replicated (FDR<0.05), findings were consistent with limited evidence for a causal effect of peripheral blood-derived DNA methylation on lung cancer (__Figure \@ref(fig:sup-fig1-07)__).

```{r sup-tab5-07, message=FALSE, results="asis"}
```

(ref:sup-fig1-07-cap) __Comparison of two-sample Mendelian randomization results when using the discovery (ARIES, n = 1018) and replication (NSHDS, n = 468)__. On the left-hand side of each column the bracketed numbers represent the number of instrumental variables for that CpG site. 

(ref:sup-fig1-07-scap) Comparison of two-sample Mendelian randomization results when using the discovery (ARIES, n = 1018) and replication (NSHDS, n = 468)

```{r sup-fig1-07, message=FALSE, results="asis", fig.cap='(ref:sup-fig1-07-cap)', fig.scap='(ref:sup-fig1-07-scap)'}
```

There was little evidence of different effect estimates between ever and never smokers at individual CpG sites (__Figure \@ref(fig:sup-fig2-07)__, Z-test for difference, P>0.5). There was some evidence for a possible effect of methylation at cg21566642-_ALPPL2_ and cg23771366-_PRSS23_ on squamous cell lung cancer (OR=0.85 [95% CI=0.75,0.97] and 0.91 [95% CI=0.84,1.00] per SD [14.4% and 5.8%] increase, respectively) as well as methylation at cg23387569-_AGAP2_, cg16823042-_AGAP2_, and cg01901332-_ARRB1_ on lung adenocarcinoma (OR=0.86 [95% CI=0.77,0.96], 0.84 [95% CI=0.74,0.95], and 0.89 [95% CI=0.80,1.00] per SD [9.47%, 8.35%, and 8.91%] increase, respectively). However, none of the results withstood multiple testing correction (FDR<0.05) (__Figure \@ref(fig:sup-fig3-07)__). For those CpGs where multiple mQTLs were used as instruments (cg05575921-_AHRR_ and cg01901332-_ARRB1_), there was limited evidence for heterogeneity in MR effect estimates (Q-test, P>0.05, __Table \@ref(tab:sup-tab6-07)__).

(ref:sup-fig2-07-cap) __DNA methylation – lung cancer Mendelian randomization effect estimates in ever and never smokers__. On the left-hand side of each column the bracketed numbers represent the number of instrumental variables for that CpG site. * indicates that the SNP(s) being used to instrument that CpG site are more than 1MB away from the CpG site in the genome (trans).

(ref:sup-fig2-07-scap) DNA methylation – lung cancer Mendelian randomization effect estimates in ever and never smokers

```{r sup-fig2-07, message=FALSE, results="asis", fig.cap='(ref:sup-fig2-07-cap)', fig.scap='(ref:sup-fig2-07-scap)'}
```

(ref:sup-fig3-07-cap) __Mendelian randomization of DNA methylation on three lung cancer subgroups__ On the left-hand side of each column the bracketed numbers represent the number of instrumental variables for that CpG site. * indicates that the SNP(s) being used to instrument that CpG site are more than 1MB away from the CpG site in the genome (trans). Squamous = squamous cell carcinoma, Small = small cell carcinoma, Adeno = adenocarcinoma.

(ref:sup-fig3-07-scap) Mendelian randomization of DNA methylation on three lung cancer subgroups

```{r sup-fig3-07, message=FALSE, results="asis", fig.cap='(ref:sup-fig3-07-cap)', fig.scap='(ref:sup-fig3-07-scap)'}
```

```{r sup-tab6-07, message=FALSE, results="asis"}
```

\pagebreak

Single mQTLs for cg05575921-_AHRR_, cg27241845-_ALPPL2_, and cg26963277-_KCNQ1_ showed some evidence of association with smoking cessation (former vs. current smokers), although these associations were not below the FDR<0.05 threshold (__Figure \@ref(fig:sup-fig4-07)__).

(ref:sup-fig4-07-cap) __Associations of mQTLs and smoking behaviours__. Some SNPs that were genotyped in the TRICL consortium were not within the TAG consortium, thus were not available for analysis here. Units for the traits: AoI (age of smoking initiation) = log years, EvN (ever vs. never smoked) = log odds, FvS (former vs current smoker) = log odds. CpD = cigarettes smoked per day.

(ref:sup-fig4-07-scap) Associations of mQTLs and smoking behaviours

```{r sup-fig4-07, message=FALSE, results="asis", fig.cap='(ref:sup-fig4-07-cap)', fig.scap='(ref:sup-fig4-07-scap)'}
```

#### Potential causal effect of _AHRR_ methylation on lung cancer risk: one sample MR {#ahrr-one-sample-mr}
Dr Rebecca Richmond completed the analysis and writing of this section.

In the CCHS, a per (average methylation-increasing) allele change in a four-mQTL allele score was associated with a 0.73% [95% CI=0.56,0.90] increase in methylation (P<1x10^-10^) and explained 0.8% of the variance in cg05575921-_AHRR_ methylation (F-statistic=74.2). Confounding factors were not strongly associated with the genotypes in this cohort (P>=0.11) (__Table \@ref(tab:sup-tab7-07)__).  Results provided some evidence for an effect of cg05575921 methylation on total lung cancer risk (HR=0.30 [95% CI=0.10,1.00] per SD (9.2%) increase) (__Table \@ref(tab:sup-tab8-07)__). The effect estimate did not change substantively when stratified by smoking status (__Table \@ref(tab:sup-tab8-07)__). \linebreak

```{r sup-tab7-07, message=FALSE, results="asis"}
```

```{r sup-tab8-07, message=FALSE, results="asis"}
```

\linebreak

Given contrasting findings with the main MR analysis, where cg05575921-_AHRR_ methylation was not causally implicated in lung cancer, and the lower power in the one-sample analysis to detect an effect of equivalent size to the observational results (power = 19% at alpha = 0.05), a further two-sample MR was performed based on the four mQTLs using data from both CCHS (sample one) and the TRICL-ILCCO consortium (sample two). Results showed no strong evidence for a causal effect of DNA methylation on total lung cancer risk (OR=1.00 [95% CI=0.83,1.10] per SD increase) (__Figure \@ref(fig:sup-fig5-07)__). There was also limited evidence for an effect of cg05575921-_AHRR_ methylation when stratified by cancer subtype and smoking status (__Figure \@ref(fig:sup-fig5-07)__) and no strong evidence for heterogeneity of the mQTL effects (__Table \@ref(tab:sup-tab9-07)__). Conclusions were consistent when MR-Egger [@Bowden2015] was applied (__Figure \@ref(fig:sup-fig5-07)__) and when accounting for correlation structure between the mQTLs  (__Table \@ref(tab:sup-tab9-07)__). 

(ref:sup-fig5-07-cap) __Two-sample Mendelian randomization analysis of DNA methylation at _AHRR_ on lung cancer__. The top four plots assess the association of DNA methylation at _AHRR_ and different lung cancer subtypes. The bottom two plots assess the association between DNA methylation and all lung cancer after stratifying into groups of ever and never smokers.

(ref:sup-fig5-07-scap) Two-sample Mendelian randomization analysis of DNA methylation at _AHRR_ on lung cancer

```{r sup-fig5-07, message=FALSE, results="asis", fig.cap='(ref:sup-fig5-07-cap)', fig.scap='(ref:sup-fig5-07-scap)'}
```

```{r sup-tab9-07, message=FALSE, results="asis"}
```

### Tumour and adjacent normal lung tissue methylation patterns {#lc-heathly-v-normal}
For cg05575921-_AHRR_, there was no strong evidence for differential methylation between adenocarcinoma tissue and adjacent healthy tissue (P=0.963), and weak evidence for hypermethylation in squamous cell carcinoma tissue (P=0.035) (__Figure \@ref(fig:fig4-07)__, __Table \@ref(tab:sup-tab10-07)__). For the other CpG sites there was evidence for a difference in DNA methylation between tumour and healthy adjacent tissue at several sites in both adenocarcinoma and squamous cell carcinoma, with consistent differences for CpG sites in _ALPPL2_ (cg2156642, cg05951221 and cg01940273), as well as cg23771366-_PRSS23_, cg26963277-_KCNQ1_, cg09935388-_GFI1_, cg0101332-_ARRB1_, cg08709672-_AVPR1B_ and cg25305703-_CASC21_. However, hypermethylation in tumour tissue was found for the majority of these sites, which is the opposite to what was observed in the EWAS analysis.

(ref:fig4-07-cap) __Differential DNA methylation in lung cancer tissue__. A comparison of methylation at each of the 16 CpG sites identified in the meta-analysis was made between lung cancer tissue and adjacent healthy lung tissue for patients with lung adenocarcinoma (__A__) and squamous cell lung cancer (__B__). Data from The Cancer Genome Atlas was used for this analysis.

(ref:fig4-07-scap) Differential DNA methylation in lung cancer tissue

\blandscape

```{r fig4-07, message=FALSE, results="asis", fig.show = "hold", out.width = "50%", fig.align = "default", fig.cap='(ref:fig4-07-cap)', fig.scap='(ref:fig4-07-scap)', fig.pos="htbp"}
```

\elandscape

```{r sup-tab10-07, message=FALSE, results="asis"}
```

### Gene expression associated with mQTLs in blood and lung tissue
Of the 10 genes annotated to the 14 CpG sites, eight genes were expressed sufficiently to be detected in lung (_AVPR1B_ and _CASC21_ were not) and seven in blood (_AVPR1B_, _CASC21_ and _ALPPL2_ were not). Of these, gene expression of _ARRB1_ could not be investigated as the mQTLs in that region were not present in the GTEx data. rs3748971 and rs878481, mQTLs for cg21566642 and cg05951221 respectively, were associated with increased expression of _ALPPL2_ in lung tissue (P=0.002 and P=0.0001). No other mQTLs were associated with expression of the annotated gene at a Bonferroni corrected P value threshold (P<0.05/19=0.0026) (__Table \@ref(tab:sup-tab11-07)__).

```{r sup-tab11-07, message=FALSE, results="asis"}
```

## Discussion {#discussion-07}
In this study, 16 CpG sites associated, at P<1x10^-7^, with lung cancer in a meta-analysis of EWAS, of which 14 have been previously identified in relation to smoke exposure [@Joehanes2016] and six were highlighted in a previous study as being associated with lung cancer [@Baglietto2017]. This previous study used the same data from the four cohorts investigated here, but in a discovery and replication, rather than meta-analysis framework. Overall, using MR we found limited evidence supporting a potential causal effect of methylation at the CpG sites identified in peripheral blood on lung cancer. These findings are in contrast to previous analyses suggesting that methylation at two CpG sites investigated (in _AHRR_ and _F2RL3_) mediated > 30% of the effect of smoking on lung cancer risk [@Fasanelli2015]. This previous study used methods which are sensitive to residual confounding and measurement error that may have biased results [@Richmond2016; @Hemani2017]. These limitations are largely overcome using MR [@Richmond2016].  While there was some evidence for an effect of methylation at some of the other CpG sites on risk of subtypes of lung cancer, these effects were not robust to multiple testing correction and were not validated in the analysis of tumour and adjacent normal lung tissue methylation nor in gene expression analysis.

A major strength of the study was the use of two-sample MR to integrate an extensive epigenetic resource and summary data from a large lung cancer GWAS to appraise causality of observational associations with >99% power. Evidence against the observational findings were also acquired through tissue-specific DNA methylation and gene expression analyses.

Limitations include potential “winner’s curse” which may bias causal estimates in a two-sample MR analysis towards the null if the discovery sample for identifying genetic instruments is used as the first sample, as was done for the main MR analysis using data from ARIES [@Burgess2011]. However, findings were similar when using replicated mQTLs in NSHDS, indicating the potential impact of this bias was minimal (__Figure \@ref(fig:sup-fig1-07)__). Another limitation relates to the potential issue of consistency and validity of the instruments across the two samples. For a minority of the mQTL-CpG associations (4 out of 16), there was limited replication across time points and in particular, 6 mQTLs were not strongly associated with DNA methylation in adults. Further, the primary data used for the first sample in the two-sample MR was ARIES, which contained no male adults at the time of the study.. If the mQTLs identified vary by sex and time, then this could bias the results. However, the replication cohort NSHDS contains adult males. Therefore, the 10 mQTLs that replicated in NSHDS are unlikely to be biased by the sex discordance. Also, the findings for cg05575921 _AHRR_ in CCHS, which contains both adult males and females, were replicated in a two-sample MR analysis, suggesting these results are also not influenced by sex discordance. Caution is therefore warranted when interpreting the null results for the two-sample MR estimates for the CpG sites for which mQTLs were not repliacted, which could be the result of weak-instrument bias.

The lack of independent mQTLs for each CpG site did not allow us to properly appraise horizontal pleiotropy in the MR analyses. Where possible I only included cis-acting mQTLs to minimise pleiotropy and investigated heterogeneity where there were multiple independent mQTLs. Three mQTLs were nominally associated with smoking phenotypes, but not to the extent that this would bias the MR results substantially. Since this study was conducted, GoDMC has released a large set of SNP-DNA methylation associations from a sample size that is over ten times that of ARIES [@Min2020]. Replication of this analysis using mQTL estimates from GoDMC may mitigate some of these limitations. 

Some of the mQTLs used influence multiple CpGs in the same region, suggesting genomic control of methylation at a regional rather than single CpG level. This was untested, but methods to detect differentially methylated regions (DMRs) and identify genetic variants which proxy for them may be fruitful in probing the effect of methylation across gene regions.

A further limitation relates to the inconsistency in effect estimates between the one- and two-sample MR analysis to appraise the causal role of _AHRR_ methylation. While findings in CCHS were supportive of a causal effect of _AHRR_ methylation on lung cancer (HR=0.30 [95% CI=0.10,1.00] per SD), in two-sample MR this site was not causally implicated (OR=1.00 [95% CI=0.83,1.10] per SD increase). It was verified that this was not due to differences in the genetic instruments used, nor due to issues of weak instrument bias. Given the CCHS one-sample MR had little power (19% at alpha = 0.05) to detect a causal effect with a size equivalent to that of the observational analysis, there should be more confidence in the results from the two-sample approach.

Peripheral blood may not be the ideal tissue to assess the association between DNA methylation and lung cancer. While a high degree of concordance in mQTLs has been observed across lung tissue, skin and peripheral blood DNA [@Shi2014], this was not directly evaluated here. A possible explanation for a lack of causal effect at _AHRR_ is due to the limitation of tissue specificity as it was found that the mQTLs used to instrument cg05575921 were not strongly related to expression of _AHRR_ in lung tissue. However, findings from MR analysis were corroborated by the lack of evidence for differential methylation at _AHRR_ between lung adenocarcinoma tissue and adjacent healthy tissue, and weak evidence for hypermethylation (opposite to the expected direction) in squamous cell lung cancer tissue. This result may be interesting in itself as smoking is hypothesized to influence squamous cell carcinoma more than adenocarcinoma. However, the result conflicts with that found in the MR analysis. Furthermore, another study investigating tumorous lung tissue (N=511) found only weak evidence for an association between smoking and cg05575921 _AHRR_ methylation, that did not survive multiple testing correction (P=0.02) [@Freeman2016]. These results do not fully exclude _AHRR_ from involvement in the disease process. _AHRR_ and _AHR_ form a regulatory feedback loop, which means that the actual effect of differential methylation or differential expression of _AHR_/_AHRR_ on pathway activity is complex [@Chen2017]. In addition, some of the CpG sites identified in the EWAS were found to be differentially methylated in the tumour and adjacent normal lung tissue comparison. While this could represent a false negative result of the MR analysis, it is of interest that differential methylation in the tissue comparison analysis was typically in the opposite direction to that observed in the EWAS. Furthermore, while this method can be used to minimize confounding, it does not fully eliminate the possibility of bias due to reverse causation (whereby cancer induces changes in DNA methylation) or intra-individual confounding e.g. by gene expression. Therefore, it doesn’t give conclusive evidence that DNA methylation changes at these sites are not relevant to the development of lung cancer.

While DNA methylation in peripheral blood may be predictive of lung cancer risk, according to the present analysis it is unlikely to play a causal role in lung carcinogenesis at the CpG sites investigated. Findings from this study issue caution over the use of traditional mediation analyses to implicate intermediate biomarkers (such as DNA methylation) in pathways linking an exposure with disease, given the potential for residual confounding in this context [@Richmond2016]. However, the findings of this study do not preclude the possibility that other DNA methylation changes (i.e. changes at different sites in the genome or in different tissues) are causally related to lung cancer (or other smoking-associated disease) [@Gao2016].

